Data Preparation

### 2019-03-12 Data Preparation.Rmd
split long chunks into smaller chunks for clarity

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache.path = "Cache/Data Preparation/")
```

```{r libraries, message = FALSE}
library(tidyverse)
library(tidytext)
library(gsheet) # import spreadsheets from google drive
library(tm)
library(hunspell) # Stemmer
```

@@

## Setup
First, we import the spreadsheet with information pertraining the Aims of the Degree (AoD) and Assessments from the drive and save it under `lists_brut`. ` lists_brut` contains 4 columns, under which we find the `19` types of assessment, the `18` aims of the degree (AoD) of the degree, and  two columns containing binary vectors indicating which assessment types and AoD  we will consider when ploting the data[^1].
Then we create a list with this same columns, but instead of having binary vectors for the plots, we keep vectors of only the names of relevant assessments and AoDs for the plots (`Assessment_plot`adnd `AoD_plot` respectively). We also any imported emtpy cells.
```{r import list}
list_AoD_assessment <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1soRA1u5zf9oLNirGmZ9yZ7m2ccPa3XFemnxkj5AVRXo/edit#gid=1239912347') %>%
  map(na.omit)

@@

```{r import d_course d_assessment and d_ILO, cache = TRUE}
d_course <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1soRA1u5zf9oLNirGmZ9yZ7m2ccPa3XFemnxkj5AVRXo/edit#gid=1655700848') %>%
  print

d_assessment <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1soRA1u5zf9oLNirGmZ9yZ7m2ccPa3XFemnxkj5AVRXo/edit#gid=1102665750') %>%
  mutate_if(
    .predicate = is.numeric,
    as.logical
    ) %>%

  # Transform data into tidy format to facilitate manipulation
  gather(
    key   = Assessment, 
    value = assessment_covered,
    Paper : Participation
    ) %>%
  filter(
    assessment_covered
    ) %>%
  select(
    - c(assessment_covered, `Comment on Assessment`)
    ) %>%
  print

d_ILO <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1soRA1u5zf9oLNirGmZ9yZ7m2ccPa3XFemnxkj5AVRXo/edit#gid=1896457050')

d_AoD_from_ILO <- d_ILO %>%
  rename(
    ILO = Objectives
    ) %>%
  mutate_if(
    .predicate = is.numeric,
    as.logical
    ) %>%

@@

## Textual Data
### Course Catalogues
In order to conduct a preliminary topic modeling of course content, we first analyze their description in the course catalogues. The corpus `corpus_catalogues` contains the pdfs of the `5` most recent course catalogues.
```{r import corpus_catalogues, cache = TRUE}
# helper function
read_in_pdfs <- function(file){

@@

### Course Manuals
To expand our topic modeling of course content, we analyse the material in the course manuals. The `corpus_manuals` contains the pdfs of the course manuals for the year 2017-2018 (most recently, available).

```{r import corpus_manuals, cache = TRUE}
extract_code <- function(string) string %>% str_sub(start = 1, end = 7) # helper functions

@@

```{r import d_transcript, cache = TRUE}
col_parsing <- cols(
  `Student Number`                      = "c",
  `Program (Abbreviation)`              = "c",
  `Appraisal Status`                    = "c",
  `Module Booking Reason (Description)` = "c",
  `Object name`                         = "c",
  `Start date`                          = "c",
  `End Date`                            = "c"
  )

@@

d_transcript <- list("Input/Raw Grades/grades1.csv", "Input/Raw Grades/grades2.csv") %>%
  map(read_csv_own_parsing) %>%
  bind_rows

@@

## Textual Data
### Extracting Text
In order to conduct a topic analysis of course content, we extract the *overview* and *description* of each course from the course catalogues, as well as the  *full text*  from the Course Manuals.

#### Extractin Text: Course Catalogues
(For this section we use:`corpus_catalogues` contains the pdfs of the `5` most recent course catalogues.)

##### Extracting Overviews

@@

```{r extracting overview, cache = TRUE}
#
# Set up

collapse_or <- function(string) string %>% str_c(collapse = "|")

years <- paste0(2014:2018, "-", 2015:2019)

@@

#
# helper functions
is_start_overview_section  <- function(string) string %>% str_detect(pattern = "^Core Courses \\(COR\\)")
is_end_overview_section    <- function(string) string %>% str_detect(pattern = "Appendix")
is_header                  <- function(string) string %>% str_detect(pattern = headers)

@@

#
# main function
extract_overview <- function(catalogue){

@@

d_text$overview <- d_text$catalogues %>%

  # extract the individual course overviews for each catalogue
  map(extract_overview) %>%

  # add the year to each observation to keep track of the evolution of the catalogue over time
  map2(
    .y = years,
    .f = function(x, year) x %>% mutate(year = year)
    ) %>%

@@

d_text <- d_text[c("overview", "manuals")]

@@

##### Extracting Descriptions

Since the overviews contain a lot of information that does not interest us, we extract the description section from the overviews. The description section contains a brief description (200-500 words) of the content of the course. To accomplish this, for each overview, we need to identify the line where the description starts and the line where it ends. The description section is always preceeded by a header saying *Description* or *Course Description* (`start_descrip`), making it relatively easy to find the starting line of the decription section with `grep`. The description section is usually followed by the literature section which starts with header saying *Literature*, *Recommended Literature* or similar (see `end_descrip`). Using `grep` together with `end_descrip`, we can again identify the ending line of the description section of almost all overviews. A handful overviews do not contain a *Literature* section. For these, the section *Instructional Format* marks the end of the description section.

```{r extracting description, eval = FALSE}

#
# Setup
symbol_to_line <- c("\r\n", "\n") %>% collapse_or
start_descrip  <- c("^Description", "^Course Description") %>% collapse_or
end_descrip    <- c(
  "^Recommended Literature", "^ ?Required Litera", "^ ?Literature$", "Literature \\(all",
  "^Instructional format$"
  ) %>% collapse_or

#
# helper functions
split              <- function(string) string %>% str_split(pattern = "\r\n|\n") %>% .[[1]]

is_start_descrip   <- function(x) x %>% str_detect(pattern = start_descrip           )
is_end_descrip     <- function(x) x %>% str_detect(pattern = end_descrip             )

#
# main function
extract_description <- function(overview){

@@

  mutate(
    is_start_descrip = line %>% is_start_descrip,
    is_end_descrip   = line %>% is_end_descrip # several ends (matching literature and instructional format) is ok because we choose the first (cumany)
  ) %>%

@@

d_text$description <- d_text$overview %>%
  mutate(
    text = text %>% map_chr(possibly(extract_description, "ERROR"))
    )
```

#### Extracting Text: Course Manuals
(For this section we use:`corpus_manuals`-contains the pdfs of the course manuals for the year 2017-2018 (most recently, available))

Since we want to use tools from dplyr and tidytext we must first transform our corpus into a dataframe.

### Tidy Text Format

To put everything into tidy text format, we first create three tibbles `d_overview_tidy`, `d_description_tidy`, `d_manual` that respectively store the overviews, descriptions and text from manuals in the tidy text format, with one row per course-year-word (and course-word for `d_manual`).
```{r tidy d_description}
# helper function
tidy_text <- function(df){

@@

d_text <- d_text %>% map(tidy_text) %>% print

@@

We first create a function `stem_hunspell` which, given a term, returns its most basic stem (the last one from the list of stem returned by `hunspell_stem`). We would like to apply `stem_hunspell` on the words of `d_overview_tidy` and `d_manual`, since they have similar structures, we row bind them into `dictionary` to ease the application of `stem_hunspell`. However, since, `stem_hunspell` is not a vectorized function and the number of words in `dictionary` is large, we first use distinct on the `word` variable to find a list containing *once* each term present in the course overviews and manuals. Then, we use `mutate` to apply the function `stem_hunspell` on each word from our dictionary and save its results as a new column `word_stem`.
```{r dictionary, cache = TRUE}
# helper function
stem_hunspell <- function(term) {

@@

dictionary <- d_text$overview$word %>%
  union(
    d_text$manuals$word
    ) %>%
  tibble(
    word = .
    ) %>%
  mutate(
    word_stem = word %>% map_chr(stem_hunspell)
    )

# TODO: "destem" terms by hand: cuss => discuss
filter(dictionary, word_stem == "cuss")

# terms for which the stem differs from the original word
filter(dictionary, word != word_stem)
```
Finally, we create a function `stem_with_dictionary` that performs a left join on the dataframe it takes as input with `dictionary`, thus adding the word stems to the original dataframe.

Then, we use `stem_with_dictionary` to include the stems of all words in `d_description`, `d_overview`, and `d_manual`.
```{r stemming data sets}
# helper function
stem_with_dictionary <- function(data){

  data %>%
    left_join(
      dictionary,
      by = "word"
      ) %>%
    rename(
      word_original = word,
      word          = word_stem
      )

d_text <- d_text %>% map(stem_with_dictionary) %>% print

@@

### Removing Stopwords
Finally, we want to filter out some uninformative words from our textual data. For this, we first store all the unwanted words in a vector `sw_owm`. We also create function `remove_sw` which receives a dataframe as input and filters out all words in `sw_own`. Then we apply this function to `d_description`, `d_overview`, and `d_manual`.
```{r}
# Own stop words
sw_own <- tibble(
    word = c(
      as.character(1 : 1e3)
      )
    )

remove_sw <- function(data){

  data %>%
    anti_join(
      stop_words,
      by = "word"
      ) %>%
    anti_join(
      sw_own,
      by = "word"
      ) %>%

    # remove words of 1 or 2 letters
    filter(
      nchar(word) > 2
      ) %>%

@@

d_text <- d_text %>% map(remove_sw) %>% print

@@

```{r}
#
# Set up
UCM_course <- d_text$overview$`Course ID` %>%
  unique %>% 
  paste(collapse = "|")

# helper function
is_offered_at_UCM <- function(x) x %>% str_detect(UCM_course)
```

```{r}
d_transcript <- d_transcript %>%

  # for simplicity, we only keep *UCM courses* taken in the framework of the *BA Liberal Arts and Sciences (UCM)* (7501)
  filter(
    `Module (Abbrev.)` %>% is_offered_at_UCM,
    `Program (Abbreviation)` == "7501"
  ) %>%

  # The dataset has multiple rows for each student - course. Following guidelines of Richard Vos, we only consider: `Appraisal (Description)` == "Grade supervisor" and `Appraisal Type` == "7055"
  filter(
    `Appraisal (Description)` == "Grade supervisor",
    `Appraisal Type`          == "7055" # removes ~ 15 observations
  )
```

```{r}
d_transcript <- d_transcript %>%

  # Remove variables with only one value
  discard(
    function(x) n_distinct(x) == 1
  ) %>%

  # Select: Student ID, course ID, when course taken and grade 
  select(
    `Student ID`   = `Student Number`,
    `Course ID`    = `Module (Abbrev.)`,
    Year_numerical = `Academic Year`,
    Period         = `Academic Session`,
    Grade          = `Grade symbol`
  )
```

```{r}

d_transcript <- d_transcript %>%

  # Clean grade variable
  mutate(

    Grade = if_else(is.na(Grade) | Grade == "NG", "0", Grade) %>% 
      str_replace(",", ".") %>% 
      as.numeric

  ) %>%

  # Clean Time variables
  mutate(

    Period = case_when(
      Period == 1   ~ "1 to 6"  ,
      Period == 2   ~ "1 to 3"  ,
      Period == 3   ~ "4 to 6"  ,
      between(Period, 100, 199) ~ "1",
      between(Period, 200, 299) ~ "2",
      between(Period, 300, 399) ~ "3",
      between(Period, 400, 499) ~ "4",
      between(Period, 500, 599) ~ "5",
      between(Period, 600, 660) ~ "6"
    ),

    time = str_c(Year_numerical, substr(Period, 1, 1), sep = ""),

    `Academic Year`= str_c(Year_numerical, Year_numerical + 1, sep = "-")

@@

```{r}
d_transcript %>%

  # highlight issue: there are several rows with unique student ID - course ID - time combination.
  add_count(
    `Student ID`, `Course ID`, time,
    sort = TRUE
    ) %>%
  print
```

>>>>

```{r setup, include = FALSE}
knitr::opts_chunk$set(cache.path = "Cache/Data Preparation/")

library(tidyverse)
library(tidytext)
library(gsheet) # import spreadsheets from google drive
library(tm)
library(hunspell) # Stemmer
```

@@

## List of AoD's and Assessments
First, we import the spreadsheet with information pertraining the Aims of the Degree (AoD) and Assessments from the drive and save it under `lists_brut`. ` lists_brut` contains 4 columns, under which we find the `19` types of assessment, the `18` aims of the degree (AoD) of the degree, and  two columns containing binary vectors indicating which assessment types and AoD  we will consider when ploting the data[^1].
Then we create a list with this same columns, but instead of having binary vectors for the plots, we keep vectors of only the names of relevant assessments and AoDs for the plots (`Assessment_plot`adnd `AoD_plot` respectively). We also any imported emtpy cells.
```{r list_AoD_assessment}
list_AoD_assessment <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1soRA1u5zf9oLNirGmZ9yZ7m2ccPa3XFemnxkj5AVRXo/edit#gid=1239912347') %>%
  map(na.omit)

@@

```{r d_course}
d_course <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1soRA1u5zf9oLNirGmZ9yZ7m2ccPa3XFemnxkj5AVRXo/edit#gid=1655700848') %>% print
```

```{r d_assessment}
d_assessment <- gsheet2tbl('https://docs.google.com/spreadsheets/d/1soRA1u5zf9oLNirGmZ9yZ7m2ccPa3XFemnxkj5AVRXo/edit#gid=1102665750') %>%
  mutate_if(
    .predicate = is.numeric,
    .funs      = as.logical
    ) %>%

  # Transform data into tidy format to facilitate manipulation
  gather(
    key   = Assessment, 
    value = assessment_covered,
    Paper : Participation
    ) %>%
  filter(
    assessment_covered
    ) %>%
  select(
    - c(assessment_covered, `Comment on Assessment`)
    ) %>%
  print
```

```{r d_AoD_from_ILO}
d_AoD_from_ILO <- d_ILO %>%
  rename(
    ILO = Objectives
    ) %>%
  mutate_if(
    .predicate = is.numeric,
    .funs      = as.logical
    ) %>%

@@

## Textual Data
In order to conduct a preliminary topic modeling of course content, we first analyze their description in the course catalogues. The corpus `corpus_catalogues` contains the pdfs of the `5` most recent course catalogues.
```{r catalogues, cache = TRUE}
# helper function
read_in_pdfs <- function(file){

@@

To expand our topic modeling of course content, we analyse the material in the course manuals. The `corpus_manuals` contains the pdfs of the course manuals for the year 2017-2018 (most recently, available).

```{r manuals, cache = TRUE}
# helper functions
extract_code <- function(string) string %>% str_sub(start = 1, end = 7)

@@

```{r d_transcript, cache = TRUE}
col_parsing <- cols(
  `Student Number`                      = "c",
  `Program (Abbreviation)`              = "c",
  `Appraisal Status`                    = "c",
  `Module Booking Reason (Description)` = "c",
  `Object name`                         = "c",
  `Start date`                          = "c",
  `End Date`                            = "c"
  )

@@

d_transcript <- paste0("Input/Raw Grades/grades", 1:2, ".csv") %>%
  map(read_csv_own_parsing) %>%
  bind_rows

@@

## Textual Data
### Extracting Overviews
In order to conduct a topic analysis of course content, we extract the *overview* and *description* of each course from the course catalogues, as well as the  *full text*  from the Course Manuals.

(For this section we use:`corpus_catalogues` contains the pdfs of the `5` most recent course catalogues.)

@@

```{r overview set up}
# Set up
collapse_or <- function(string) string %>% str_c(collapse = "|")

@@

years <- paste0(2014:2018, "-", 2015:2019)
```

```{r overview helper functions}
# helper functions in order of use
is_start_overview_section  <- function(string) string %>% str_detect(pattern = "^Core Courses \\(COR\\)")
is_end_overview_section    <- function(string) string %>% str_detect(pattern = "Appendix")
is_header                  <- function(string) string %>% str_detect(pattern = headers)

@@

add_year <- function(x, yr) x %>% mutate(year = yr)
```

```{r extract_overview}
extract_overview <- function(catalogue){

@@

```{r overview, cache = TRUE}
d_text$overview <- d_text$catalogues %>%

  map(extract_overview) %>%

  # add the year to each observation to keep track of the evolution of the catalogue over time
  map2(years, add_year) %>%

@@

# only keep the overviews and manuals for clarity
d_text <- d_text[c("overview", "manuals")]

@@

### Extracting Descriptions
Since the overviews contain a lot of information that does not interest us, we extract the description section from the overviews. The description section contains a brief description (200-500 words) of the content of the course. To accomplish this, for each overview, we need to identify the line where the description starts and the line where it ends. The description section is always preceeded by a header saying *Description* or *Course Description* (`start_descrip`), making it relatively easy to find the starting line of the decription section with `grep`. The description section is usually followed by the literature section which starts with header saying *Literature*, *Recommended Literature* or similar (see `end_descrip`). Using `grep` together with `end_descrip`, we can again identify the ending line of the description section of almost all overviews. A handful overviews do not contain a *Literature* section. For these, the section *Instructional Format* marks the end of the description section.

```{r description set up}
# Setup
symbol_to_line <- c("\r\n", "\n") %>% collapse_or
start_descrip  <- c("^Description", "^Course Description") %>% collapse_or
end_descrip    <- c(
  "^Recommended Literature", "^ ?Required Litera", "^ ?Literature$", "Literature \\(all",
  "^Instructional format$"
  ) %>% collapse_or
```

```{r description helper function}
# helper functions
split              <- function(string) string %>% str_split (pattern = symbol_to_line) %>% .[[1]]

is_start_descrip   <- function(string) string %>% str_detect(pattern = start_descrip )
is_end_descrip     <- function(string) string %>% str_detect(pattern = end_descrip   )
```

```{r extract_description, eval = FALSE}
extract_description <- function(overview){

@@

    mutate(
      is_start_descrip = line %>% is_start_descrip,
      is_end_descrip   = line %>% is_end_descrip # Having several ends (e.g. "literature" and "instructional format") is ok because we only use the first (cumany)
      ) %>%

@@

```{r description}
d_text$description <- d_text$overview %>%
  mutate(text = text %>% map_chr(extract_description))
```

### Tidy Text Format
To put everything into tidy text format, we first create three tibbles `d_overview_tidy`, `d_description_tidy`, `d_manual` that respectively store the overviews, descriptions and text from manuals in the tidy text format, with one row per course-year-word (and course-word for `d_manual`).
```{r tidy_text}
# helper function
tidy_text <- function(df){

@@

```{r tidy d_text}
d_text <- d_text %>% map(tidy_text) %>% print

@@

We first create a function `stem_hunspell` which, given a term, returns its most basic stem (the last one from the list of stem returned by `hunspell_stem`). We would like to apply `stem_hunspell` on the words of `d_overview_tidy` and `d_manual`, since they have similar structures, we row bind them into `dictionary` to ease the application of `stem_hunspell`. However, since, `stem_hunspell` is not a vectorized function and the number of words in `dictionary` is large, we first use distinct on the `word` variable to find a list containing *once* each term present in the course overviews and manuals. Then, we use `mutate` to apply the function `stem_hunspell` on each word from our dictionary and save its results as a new column `word_stem`.

```{r stem}
# helper function
stem <- function(term) {

@@

```{r dictionary, cache = TRUE}
dictionary <- union(d_text$overview$word, d_text$manuals$word) %>%

  tibble(word = .) %>%

  mutate(word_stem = word %>% map_chr(stem))

# terms for which the stem differs from the original word
filter(dictionary, word != word_stem)
```

```{r destem}
# TODO: "destem" terms by hand: cuss => discuss
filter(dictionary, word_stem == "cuss")
```

Finally, we create a function `stem_with_dictionary` that performs a left join on the dataframe it takes as input with `dictionary`, thus adding the word stems to the original dataframe.

Then, we use `stem_with_dictionary` to include the stems of all words in `d_description`, `d_overview`, and `d_manual`.
```{r stem_with_dictionary}
# helper function
stem_with_dictionary <- function(data){

  data %>%

    left_join(dictionary, by = "word") %>%

    rename(
      word_original = word,
      word          = word_stem
      )

```{r stem d_text}
d_text <- d_text %>% map(stem_with_dictionary)

@@

### Removing Stopwords
Finally, we want to filter out some uninformative words from our textual data. For this, we first store all the unwanted words in a vector `sw_owm`. We also create function `remove_sw` which receives a dataframe as input and filters out all words in `sw_own`. Then we apply this function to `d_description`, `d_overview`, and `d_manual`.
```{r sw_own}
# Own stop words
sw_own <- 1 : 1e3 %>% as.character %>% tibble( word = .)
```

```{r remove_sw}
# helper function
remove_sw <- function(data){

  data %>%

    anti_join(stop_words, by = "word") %>%

    anti_join(sw_own, by = "word") %>%

    # remove words of 1 or 2 letters
    filter(nchar(word) > 2) %>%

@@

```{r sw d_text}
d_text <- d_text %>% map(remove_sw) %>% print

@@

```{r d_transcript set up}
# Set up
UCM_course <- d_text$overview$`Course ID` %>%
  unique %>% 
  paste(collapse = "|")
```

```{r d_transcript helper function}
# helper function
is_offered_at_UCM <- function(x) x %>% str_detect(UCM_course)
```

```{r d_transcript row selection}
d_transcript <- d_transcript %>%

  # for simplicity, we only keep 
  # *UCM courses*  (courses present in the UCM catalogue)
  # taken in the framework of the *BA Liberal Arts and Sciences (UCM)* (Program (Abbreviation) = 7501)

  filter(
    `Module (Abbrev.)` %>% is_offered_at_UCM,
    `Program (Abbreviation)` == "7501"
    ) %>%

  # The dataset has multiple rows for each student - course.
  # Following guidelines of Richard Vos, we only consider: 
  # `Appraisal (Description)` == "Grade supervisor"
  # and `Appraisal Type` == "7055"

  filter(
    `Appraisal (Description)` == "Grade supervisor",
    `Appraisal Type`          == "7055" # removes ~ 15 observations
    )
```

```{r d_transcript column selection}
d_transcript <- d_transcript %>%

  # Remove variables with only one value (~10)
  discard(function(x) n_distinct(x) == 1) %>%

  # Select: Student ID, course ID, when a course is taken and the grade 
  select(
    `Student ID`   = `Student Number`,
    `Course ID`    = `Module (Abbrev.)`,
    Year_numerical = `Academic Year`,
    Period         = `Academic Session`,
    Grade          = `Grade symbol`
    )
```

```{r clean_grade}
clean_grade <- function(grade){

  if(is.na(grade) | grade == "NG") grade <- "0"

  grade %>% str_replace(",", ".") %>% as.numeric

```{rclean_period}
clean_period <- function(period){

  # Course spanning over multiple period are recorder with a 1, 2 or 3
  if(period == 1) return("1 to 6")
  if(period == 2) return("1 to 3")
  if(period == 3) return("4 to 6")

  # Course spanning over a single period are recorded in the 100's, 200's, etc.
  # The digit of the hundred indicates the period.
  period %/% 100 %>% as.character

```{r clean_year}
clean_year <- function(year) str_c(year, year + 1, sep = "-")
```

```{r clean d_transcript}
d_transcript <- d_transcript %>%

  mutate(

    # Clean grade, period and year
    Grade          = Grade  %>% map_dbl(clean_grade ),
    Period         = Period %>% map_chr(clean_period),
    `Academic Year`= Year_numerical %>% map_chr(clean_year),

    # extract first period in which course spanning over mutliple periods is given
    period_numerical = substr(Period, 1, 1),

    # Combine year and period to have be able to order the course in time using a single variable
    time = str_c(Year_numerical, period_numerical)

@@

```{r issue d_transcript}
d_transcript %>%

  # highlight issue: there are several rows with unique student ID - course ID - time combination.
  add_count(
    `Student ID`, `Course ID`, time,
    sort = TRUE
    ) %>%
  print

#TODO: list issues here
```
